---
output: 
  bookdown::rtf_document2:
    global_numbering: FALSE

## Note: remember to reference figures in bookdown, you need \@ref(fig:lable) using parentheses. The label is the name of the chunk
---

# Sustainability

## The definition

The sustainability concept had a defining moment in the Brundtland Commission by
which it was coined as: *"meeting the needs of the present without compromising
the ability of future generations to meet their own needs."* This definition
frames most of the usage of the term sustainability today with some amendments.
Today's globally accepted guiding principles are captured in the united nations'
sustainable development goals, the SDG's. The SDG's particularly emphasize the
just distribution of the cost and benefits of sustainability. The concept of
sustainability, its history, and interpretation are as varied as there are
people using it.

Don't make the mistake of thinking this is a shortcoming, though! Think of
sustainability as a process, a conversation method, or a guiding terrain in the
social discourse about our common future and visions. Just as life itself,
sustainability is a living concept that requires reflection, diversity,
feedback, and experimentation. In general, the term sustainability is used at
the level of societies, nations, or the human species, referring to a collective
decision-making process. But sustainability also fundamentally applies to all
levels of living organisms from the individual to life on earth. Sustainability
is ultimately the quest for long term survival and living well.

Beyond the minimum criteria of surviving, the term sustainability also refers to
how to persist, i.e., the quality of survival, the degree of fulfillment of
needs. Today, the focus is on human well-being and the distribution of
well-being among all involved parties. Wellbeing is a problematic concept since
well-being is highly individual and scale-dependent. How do we measure
well-being, and for whom? Is well-being substitutable or absolute? How does the
environment affect well-being? Is there optimal well-being? Will humans ever be
content? What level of inequality in wellbeing is acceptable?

The term *"resilience"* is related to sustainability and has been a highly
influential concept. As used in sustainability science and practice, resilience
embeds the processes that enable any system to persist through time while
maintaining its capacity to adapt and renew itself without losing its identity.
The latter is particularly relevant to any living system or society. Resilience
is based on complex adaptive systems (CAS) perspective, as explicitly stated by
the founder of the concept, C.S. Holling. In many ways, resilience and CAS refer
to similar phenomena.

## The process

Sustainability is about making decisions so human systems' environmental impacts
do not negatively shift the balance between benefits and harms. Such a shift
will ultimately affect human well-being. The environmental system consists of
physical and biological systems. From these, the human system receives benefits
(food, water, shelter, beneficial climatic environment) and harms (diseases,
natural catastrophes, and climatically challenging conditions). The unjust
distribution between the benefits and harms is well known and has been a
priority for the SDGs to recognize and address. What we call human well-being
here is one of the more difficult concepts in sustainability science. It is
still debated what measures best can guide sustainable decisions and future
planning.

A fundamental problem in defining well-being is that there is a trade-off
between its measurability and relevance. Money is measurable but has been shown
to only partly and indirectly correlate with human well-being. Aspects that
humans value most, such as meaning, feeling loved, a predictable future, and
opportunities, are much harder to quantify. Whatever the measure we agree on,
the premise of sustainability is that any impact on the environmental system
should not reduce future well-being (we will be looking much deeper into this
later). This is a problematic criterion as the possible interactions by which
activities can give feedback into the human system through benefits and harms
are difficult to predict. This makes sustainability science essentially to be
about:

1)  understanding the consequences of activities and impacts on benefits and
    harms, and
2)  designing decision processes in which distributed knowledge, power and
    observations need to be combined efficiently to maintain experienced
    wellbeing over time.

Sustainability can be seen as a state of a system in which the human system's
maintenance does not negatively impact the environmental system's support. Or it
can be thought of as the process to develop towards this state.

```{r decisions, echo = FALSE, out.width='100%', fig.align='center', fig.cap="The human system in pink is a part of and embedded in the natural system. Activities and impacts on the natural system result in chain of interactions eventually affecting the benefits and harms that the human system experiences."}
knitr::include_graphics(path = "figures/Basic_Decisions.svg")
```

Sustainability is about making smart decisions such that the human system can
develop without the activities resulting in a change of balance between benefits
and harms to reduce future wellbeing (Fig \@ref(fig:decisions)). Since
decision-making is at the heart of sustainable development, we will look closer
at a decision's basic components. Knowledge about decision-making highlights
where and how information resides. We can also understand when and why
collective decision-making is efficient. We will also use the components of
decision-making as a general structure for this course.

Individuals make decisions, sometimes with a formal protocol such as democratic
voting. Some individuals are selected to positions of authority with decision
powers. The individual's decision will still, in part, depend on her/his
perspective and experiences. There are both positive effects and challenges with
a diverse set of perspectives in the decision process. Personal biases will
always be part of human decision-making, but there are methods to reduce their
negative effects.

Decision-making is a cognitive process by which we connect perception with
activities in the physical world. **Perception** is the sensory information that
individuals get from the environmental context. The information can also be
acquired through channels such as literature and other media. **Activities** are
the result of our cognitive process and also include physically affecting the
environment and information sharing. Perceptions from our impression of the
situation mediated by our past experiences and understanding. In this step,
individual biases are already expressed. The bias is evident, for example, when
politicians describe the same situation in very different terms.

We then come to the three pillars of decisions: options, understanding, and
objectives. **Options** depend on the assessment of the situation and an
understanding of the world. Innovation is in part about expanding the set of
options by a deeper understanding of the situation. **Understanding** is
generated by personal experiences of interacting with the world and by using
language to share experiences. It is useful to reflect on how very personal our
understanding of the world is, and what this may mean for decision making. We
will discuss this more in the section on "Perspectives." **Objectives** is what
we use to select among the outcomes expected from different options. Our
understanding forms the expected outcomes. The understanding can be either from
past experiences or through reasoning and planning. To learn from reason is a
relatively unique but not entirely exclusive trait for human beings. Thus,
**decisions** are the cognitive process of selecting among the different options
with expected outcomes and implementing these into **actions**.

```{r embedded-decisions, echo=FALSE, out.width='100%', fig.align='center', fig.cap="Conceptual representation of a general decision process. There are six main capacities that make up the process. Perception is how we create a representation of the environment we are embedded in (light green). Perception then determines what options we perceive to be available. Innovation can occur when understanding generates new options. The activity is how we affect the environment we are embedded in. Understanding and objectives guide the decision. This scheme can also be seen at the level of groups or societies, where the capacities can reside in one or several individuals."}
knitr::include_graphics(path = "figures/Embedded_Decisions.svg")
```

It is useful to have this individualistic perspective when discussing how
collective action, management, governance, and power work. The world is complex
and diverse, and the socio-ecological context shapes us. To solve a complex
problem that requires cooperation, we can use the diversity of individual
world-views to address this complexity. Figure \@ref(fig:embedded-decisions)
illustrates that the decision process is embedded in a larger system, the
environment we depend on for an individual or a group. Governments, companies,
NGOs, and individuals can contribute to different parts of this scheme and
influence each decision-making process component. For example, some act as
vigilantes, shining light on situations that need addressing. Some work
tirelessly to increase understanding among actors. Others provide the activities
needed, such as doctors without borders. Governance today is about connecting
the many parts of the social fabric, preferably under sustainable and just
objective.

Sustainability at the individual, the group, the nation, or the planet level, is
about creating an efficient and just distributed decision system. When
sustainability acts as an overarching objective, even a largely distributed
decision system such as a human society can transform its operations.
Sustainability science tries to understand our activities' consequences and give
feedback on benefits and harms through the environmental system. One important
insight in the last decades has been the importance of seeing the decisions,
management, and governance as part of the process that generates understanding.
This is called adaptive management of adaptive governance (Folke et al. 2005).

Thus to summarise, sustainability science provides:

1)  A definition of sustainability
2)  The process of achieving it through decision making
3)  The Scientific process to generate the understanding to guide the decisions
    for sustainability

If interested, more reading is found here:

Dietz, Thomas, Elinor Ostrom, and Paul C. Stern. 2003. "The Struggle to Govern
the Commons." Science 302 (5652): 1907--12.

Social Problems: Continuity and Change (2016): 15.3 The Environment. University
of Minnesota Libraries Publishing. Link to chapter:
<https://open.lib.umn.edu/socialproblems/chapter/15-3-the-environment/>

Schill, Caroline, John M. Anderies, Therese Lindahl, Carl Folke, Stephen
Polasky, Juan Camilo Cárdenas, Anne-Sophie Crépin, Marco A. Janssen, Jon
Norberg, and Maja Schlüter. 2019. "A More Dynamic Understanding of Human
Behaviour for the Anthropocene." Nature Sustainability 2 (12): 1075--82.
